Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-06-28 Chapter_1/04_Manuscript/Manuscript.tex
some typos and grammatical checking

While an explicit representation of time yields several advantages, it also
requires that step lengths and turning angles are modeled properly
\citep{Kanagaraj.2013}, so that dispersal durations between habitat patches can
be reliably approximated. Accounting for and correctly rendering this movement
kernel is the main strength of ISSFs, which is why we believe that the framework
is particularly well suitded for simulating animal dispersal and assessing
landscape connectivity. In principle, the ISSF framework allows to define an
infinite amount of movement kernels, each representing movement behavior under
different conditions \citep{Fieberg.2020}. It can therefore accomodate highly
diverse movement modes, including fast paced directional movements and slow
paced tortuous movements. Compared to models that disregard an animal's movement
kernel, this allows to model directional peristence, a movement characteristic
regularly observed in dispersing animals. To date, however, we only accounted
for 1. order autocorrelation. That is, we only considered correlations between
two subsequent steps. Modeling higher-order autocorrelation may be desirable,
yet will require more extensive data.

@@

(2) While endpoints do not need to be defined in IBMMs, source points will still
need to be provided by the modeler. Here, we randomly placed source points
within protected areas large enough to sustain viable wild dog populations,
knowing that the species primarily survives in these formally protected areas
\citep{Woodroffe.1999, DaviesMostert.2012, Woodroffe.2012, VanDerMeer.2014}. In
some cases, exact locations of potential source populations are known and source
points can be placed accordingly \citep{Kanagaraj.2013}. Moreover, if abundance
estimates are available, these can be used to inform the relative number of
dispersers initiated at each location. In other cases, comparable knowledge may
be lacking and it could be more beneficial to delineate likely source patches
based on habitat suitability models (e.g. \citealp{Squires.2013}). Either way, a
meaningful selection of source points is not a problem unique to IBMMs and
applies to any connectivity modelling technique. As highlighted by
\cite{Signer.2017} the influence of the exact location of source points
decreases as the number of simulated steps is increased, yet their goal was
assessing steady state utilization distributions and not to assess landscape
connectivity.

(3) When employing IBMMs, one also needs to decide on meaningful dispersal
durations (number of simulated steps). We decided to simulate individuals for
2'000 steps, which is at the upper end of observed dispersal durations and may
have resulted in an overestimated representation of landscape connectivity.
Instead of enforcing homogeneous dispersal durations, one could also sample
dispersal durations from on observed dispersal events. Alternatively, after each
step, individuals could be faced with a probability to settle, where the
probability is contingent on the surrounding landscape and cumulative dispersal
distance travelled. Because we only observed few dispersal events and due to the
wild dog's ability to cover vast dispersal distance \citep{DaviesMostert.2012,
Masenga.2016, Cozzi.2020} we decided not to apply this approach. Regardless of
this, we think that in most cases it will be more insightful to simulate
relatively extensive dispersal events and only subsample afterwards.

(4) Unless simulated individuals are strongly drawn towards a point of
attraction, some simulated individuals will eventually approach a map boundary
and the modeller needs to define a rule to handle such situations. One feasible
option would be to terminate the trajectory as soon as it hits a map boundary,
implicitly assuming that the simulated animal left the study area and will not
return. The downside to this approach is that source points located near map
boundaries will inevitably produce many such dead ends. Alternatively, one could
remove transgressing random steps from the set of proposed random steps, thereby
forcing virtual dispersers to bounce off virtual boundaries and remain within
the main study area. This may seem like an elegant solution, yet again, it
potentially results in substantial edge effects when source points are located
near map boundaries. Finally, one could extend the study area by an artificial
buffer zone through which dispersers are allowed to leave and re-enter the main
study area. Of course this does not fully mitigate the issue, since some
dispersers may still approach the outer map boundary, however, as
\cite{Koen.2010} have shown, such an artificial buffer can often help to
mitigate edge effects. In theoretical applications, this issue can be
circumvented by simulating movement on a torus \citep{Hodel.2021b}, yet this is
unfeasible in most real world applications.

@@

\subsection{Maps (10\%)}
The heatmap resulting from our dispersal simulation suggests that a large
portion of dispersers traverses the Moremi NP and the Chobe NP in northern
Botswana. Since we already highlighted the same area in our previous analysis
\citep{Hofmann.2021}, this result was to be expected. Nevertheless, we believe
it reinforces our notion that the area acts as crucial hub inside the KAZA-TFCA.
The very same area stands out on the betweenness map, showing that the region is
not only frequently traversed, but actually serves as stepping stone into more
remote regions of the KAZA-TFCA. As such, the area exemplifies a region were
both traversal frequency and betweenness scores are high. In contrast, the
betweenness map suggests the presence of a corridor moving from Angola into
Zambia's Kafue NP, yet when looking at the heatmap, one realizes that the area
is not used very often. Consequently, despite the corridor's importance for
linking Angola's NPs to Zambia's Np, only very few simulated dispersers
successfully traversed it. This is also reflected in the interpatch connectivity
map, where the Kafue NP only receives very few links from the central region of
the KAZA-TFCA.

Each map that we produced from simulated trajectories accentuates a different
aspect of connectivity. The heatmap, for example, puts emphasis on areas where
movement is concentrated, regardless whether such areas represent ``dead ends''
that do not necessarily connect distinct patches. Therefore, the heatmap is
useful to determine locations where a disperser are likely to move to, yet not
necessarily to uncover corridors that link valuable habitats. The betweenness
map, on the other hand, brings out those areas that are relevant in connecting
different regions in the landscape. In this regard, the betweenness map is most
pertinent to the delineation of dispersal corridors and may serve as an
alternative to least-cost algorithms. Finally, the inter-patch connectivity map
illustrates the frequency at which dispersal between distinct habitat patches
occurs, as well as the average dispersal duration that passes when individuals
move between those patches. This knowledge serves to gauge the amount of
geneflow between habitat patches and can be employed to determine whether two
habitats are connected or not.

@@

% Other studies that use similar simulations
Our approach of simulating movement to assess connectivity is most closely
related to the works of \cite{Clark.2015} and \cite{Zeller.2020}, who used
regular step selection functions to model habitat preferences by black bears and
simulated movement to assess connectivity. However, neither of these studies
attempted to jointly model habitat and movement kernels, thereby precluding a
more mechanistic understanding of movement. Moreover, both studies parametrized
models using data of resident black bearsk, which may lead to a biased view on
connectivity, as dispersers may move substantially further, thereby improving
connectivity among habitat patches \citep{Elliot.2014}.

@@

% Social landscape
We have previously attributed the weak significance of distance to water to the
fact that we did not control for the presence or absence of conspecifics. We
stick to this reasoning as our expanded model still shows a rather large
uncertainty around the respective beta coefficients. To better gauge the
importance and influence of this covariate, future studies will need to control
for inter- and intra-speficic interactions that may explain why and when
dispersers are attracted to or afraid of waterbodies. \cite{Fortin.2005}, for
instance, found that elk movement was significantly impacted by the density of
wolf in the area, such that habitat preferences strictly differed depending on
the presence or absence of wolves. The decision to settle is likely related to
the presence or absence of conspecifics. Hence, the exact dispersal duration and
distance will not be independent of current wild dog densities. In dispersing
wolves, for instance, the longest dispersal distances have been observed in
low-density populations \citep{Boyd.2005, Wabakken.2007}. The dispersal duration
may thus be determined by the by the amount of isolation between subpopulations
and population densities \citep{DaviesMostert.2012}.

@@

% Conclusion
To this end, we have exemplified the use of ISSFs as a simple framework based on
which researchers can parametrize dispersers' habitat and movement preferences
and at the same time use the parametrized model to as an individual-based
movement model to simulate dispersal and examine landscape connectivity.
Furthermore, we suggest to analyse simulated trajectories using three
complementary maps, each focused on a different aspect of connectivity.
Ultimately, we hope to have sparked interest in the uprising framework of step
selection functions for investigating dispersal behavior and landscape
connectivity. Nevertheless, we do not attempt to dismiss the application of
traditional connectivity models such as least-cost methods or circuit theory by
any means. Rather, we propose to use simulations from ISSF-IBMMs as a simple but
powerful tool that complemets earlier methods in an attempt provide a more
comprehensive understanding of landscape connectivity.

>>>>

While an explicit representation of time yields several advantages, it also
requires that step lengths and turning angles are modeled properly
\citep{Kanagaraj.2013}, so that dispersal durations between habitat patches can
be reliably approximated. Accounting for and correctly rendering this movement
kernel is the main strength of ISSFs, which is why we believe that the framework
is particularly well suited for simulating animal dispersal and assessing
landscape connectivity. In principle, the ISSF framework allows to define an
infinite amount of movement kernels, each representing movement behavior under
different conditions \citep{Fieberg.2020}. It can therefore accommodate highly
diverse movement modes, including fast paced directional movements and slow
paced tortuous movements. Compared to models that disregard an animal's movement
kernel, this allows to model directional persistence, a movement characteristic
regularly observed in dispersing animals. To date, however, we only accounted
for 1. order autocorrelation. That is, we only considered correlations between
two subsequent steps. Modeling higher-order autocorrelation may be desirable,
yet will require more extensive data.

@@

(2) While endpoints do not need to be defined in IBMMs, source points will still
need to be provided by the modeler. Here, we randomly placed source points
within protected areas large enough to sustain viable wild dog populations,
knowing that the species primarily survives in these formally protected areas
\citep{Woodroffe.1999, DaviesMostert.2012, Woodroffe.2012, VanDerMeer.2014}. In
some cases, exact locations of potential source populations are known and source
points can be placed accordingly \citep{Kanagaraj.2013}. Moreover, if abundance
estimates are available, these can be used to inform the relative number of
dispersers initiated at each location. In other cases, comparable knowledge may
be lacking and it could be more beneficial to delineate likely source patches
based on habitat suitability models (e.g. \citealp{Squires.2013}). Either way, a
meaningful selection of source points is not a problem unique to IBMMs and
applies to any connectivity modeling technique. As highlighted by
\cite{Signer.2017} the influence of the exact location of source points
decreases as the number of simulated steps is increased, yet their goal was
assessing steady state utilization distributions and not to assess landscape
connectivity.

(3) When employing IBMMs, one also needs to decide on meaningful dispersal
durations (number of simulated steps). We decided to simulate individuals for
2'000 steps, which is at the upper end of observed dispersal durations and may
have resulted in an overestimated representation of landscape connectivity.
Instead of enforcing homogeneous dispersal durations, one could also sample
dispersal durations from on observed dispersal events. Alternatively, after each
step, individuals could be faced with a probability to settle, where the
probability is contingent on the surrounding landscape and cumulative dispersal
distance traveled. Because we only observed few dispersal events and due to the
wild dog's ability to cover vast dispersal distance \citep{DaviesMostert.2012,
Masenga.2016, Cozzi.2020} we decided not to apply this approach. Regardless of
this, we think that in most cases it will be more insightful to simulate
relatively extensive dispersal events and only subsample afterwards.

(4) Unless simulated individuals are strongly drawn towards a point of
attraction, some simulated individuals will eventually approach a map boundary
and the modeler needs to define a rule to handle such situations. One feasible
option would be to terminate the trajectory as soon as it hits a map boundary,
implicitly assuming that the simulated animal left the study area and will not
return. The downside to this approach is that source points located near map
boundaries will inevitably produce many such dead ends. Alternatively, one could
remove transgressing random steps from the set of proposed random steps, thereby
forcing virtual dispersers to bounce off virtual boundaries and remain within
the main study area. This may seem like an elegant solution, yet again, it
potentially results in substantial edge effects when source points are located
near map boundaries. Finally, one could extend the study area by an artificial
buffer zone through which dispersers are allowed to leave and re-enter the main
study area. Of course this does not fully mitigate the issue, since some
dispersers may still approach the outer map boundary, however, as
\cite{Koen.2010} have shown, such an artificial buffer can often help to
mitigate edge effects. In theoretical applications, this issue can be
circumvented by simulating movement on a torus \citep{Hodel.2021b}, yet this is
unfeasible in most real world applications.

@@

\subsection{Maps (10\%)}
The heatmap resulting from our dispersal simulation suggests that a large
portion of dispersers traverses the Moremi NP and the Chobe NP in northern
Botswana. Since we already highlighted the same area in our previous analysis
\citep{Hofmann.2021}, this result was to be expected. Nevertheless, we believe
it reinforces our notion that the area acts as crucial hub inside the KAZA-TFCA.
The very same area stands out on the betweenness map, showing that the region is
not only frequently traversed, but actually serves as stepping stone into more
remote regions of the KAZA-TFCA. As such, the area exemplifies a region were
both traversal frequency and betweenness scores are high. In contrast, the
betweenness map suggests the presence of a corridor moving from Angola into
Zambia's Kafue NP, yet when looking at the heatmap, one realizes that the area
is not used very often. Consequently, despite the corridor's importance for
linking Angola's NPs to Zambia's Np, only very few simulated dispersers
successfully traversed it. This is also reflected in the inter-patch
connectivity map, where the Kafue NP only receives very few links from the
central region of the KAZA-TFCA.

Each map that we produced from simulated trajectories accentuates a different
aspect of connectivity. The heatmap, for example, puts emphasis on areas where
movement is concentrated, regardless whether such areas represent ``dead ends''
that do not necessarily connect distinct patches. Therefore, the heatmap is
useful to determine locations where a disperser are likely to move to, yet not
necessarily to uncover corridors that link valuable habitats. The betweenness
map, on the other hand, brings out those areas that are relevant in connecting
different regions in the landscape. In this regard, the betweenness map is most
pertinent to the delineation of dispersal corridors and may serve as an
alternative to least-cost algorithms. Finally, the inter-patch connectivity map
illustrates the frequency at which dispersal between distinct habitat patches
occurs, as well as the average dispersal duration that passes when individuals
move between those patches. This knowledge serves to gauge the amount of
gene flow between habitat patches and can be employed to determine whether two
habitats are connected or not.

@@

% Other studies that use similar simulations
Our approach of simulating movement to assess connectivity is most closely
related to the works of \cite{Clark.2015} and \cite{Zeller.2020}, who used
regular step selection functions to model habitat preferences by black bears and
simulated movement to assess connectivity. However, neither of these studies
attempted to jointly model habitat and movement kernels, thereby precluding a
more mechanistic understanding of movement. Moreover, both studies parametrized
models using data of resident black bears, which may lead to a biased view on
connectivity, as dispersers may move substantially further, thereby improving
connectivity among habitat patches \citep{Elliot.2014}.

@@

% Social landscape
We have previously attributed the weak significance of distance to water to the
fact that we did not control for the presence or absence of conspecifics. We
stick to this reasoning as our expanded model still shows a rather large
uncertainty around the respective beta coefficients. To better gauge the
importance and influence of this covariate, future studies will need to control
for inter- and intra-specific interactions that may explain why and when
dispersers are attracted to or afraid of waterbodies. \cite{Fortin.2005}, for
instance, found that elk movement was significantly impacted by the density of
wolf in the area, such that habitat preferences strictly differed depending on
the presence or absence of wolves. The decision to settle is likely related to
the presence or absence of conspecifics. Hence, the exact dispersal duration and
distance will not be independent of current wild dog densities. In dispersing
wolves, for instance, the longest dispersal distances have been observed in
low-density populations \citep{Boyd.2005, Wabakken.2007}. The dispersal duration
may thus be determined by the by the amount of isolation between subpopulations
and population densities \citep{DaviesMostert.2012}.

@@

% Conclusion
To this end, we have exemplified the use of ISSFs as a simple framework based on
which researchers can parametrize dispersers' habitat and movement preferences
and at the same time use the parametrized model to as an individual-based
movement model to simulate dispersal and examine landscape connectivity.
Furthermore, we suggest to analyse simulated trajectories using three
complementary maps, each focused on a different aspect of connectivity.
Ultimately, we hope to have sparked interest in the uprising framework of step
selection functions for investigating dispersal behavior and landscape
connectivity. Nevertheless, we do not attempt to dismiss the application of
traditional connectivity models such as least-cost methods or circuit theory by
any means. Rather, we propose to use simulations from ISSF-IBMMs as a simple but
powerful tool that complements earlier methods in an attempt provide a more
comprehensive understanding of landscape connectivity.
